---
title: "Data Science Course Notes"
author: "Joao Pinelo Silva"
date: "September 11, 2015"
output: html_document
---
READ ME: These are my personal notes of studying R. It's a draft. A rather disorganised and incomplete draft. Use at your own peril :). If you find any mistake, please make a pull request.

### Calculating / Estimating Memory Requirements  
DATA: Data frame with 1,500,000 rows and 120 columns of numeric data and a 64 bits

Formula  
_Number of rows x Number of columns x 8 bytes_  

Why 8 bytes?  machine/OS is 64 bits = 8 bytes, since there are 8 bits per byte

1,500,000 x 120 x 8 = 1440000000  

1440000000 / 2^20 bytes/MB = 1,373.29 MB = 1.34 GB

### R Connections  
For example opening, reading and or writing files, open URL...

### Create File
``` {r, eval = FALSE}
dput(x, file = "dputed_x.R")
```
file() basic arguments:  

* description = name of file;  
* open = code indicating:  
    + "r" read only,  
    + ""w writing and initializing a new file,  
    + "a" appending,  
    + "rb", "wb", "ab" reading, writing, or appending in binary mode (windows)  

### Subsetting
* `[` returns object of same class as original - can be used to extract more than one element (one exception)  
* `[[` extract elements of list or df. Can extract a single element. Not necessarily resturn list or df.  
* `$ extract elements of list or df by name.  
  
Order: _row_, _Column_
   
   
**Subsetting vectors**  

```{r, echo = TRUE, eval = TRUE}
a <- rnorm(20)
b <- rep(NA, 20)
x <- sample(c(a,b), 15)
x

x[1:10]
x[is.na(x)]
y <- x[!is.na(x)]
y[y > 0]
x[x > 0]
x[!is.na(x) & x > 0]
x[c(3, 5, 7)]
x[c(-2, -10, -12, -15)] # same as below, putting minus before 'c'
x[-c(2, 10, 12, 15)]

vect <- c(foo = 11, bar = 2, norf = NA)
names(vect)

vect2 <- c(11, 2, NA)
names(vect2) <- c("foo", "bar", "norf")
vect2

vect["bar"]
vect[c("foo", "bar")]

```











**Subsetting lists**  

```{r, echo = TRUE, eval = TRUE}
x <- list(foo = 1:4, bar = 0.6, baz = "hello")
name <- "foo"
x[[name]] # computed index for 'foo'
x$name # element 'name' does not exist in x, therefore NULL below
x$foo # element 'foo' does exist in x, and therefore its elements are printed below
```


* `[` returns element of list. Cannot use `[[` or `$` to extract multiple elements of list.  
* `[[` returns neste elements - the element inside the list element. Can take integer seq.   

```{r, echo = TRUE, eval = TRUE}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))  
x[[c(1, 3)]]
x[[1]][[3]]
x[[c(2, 1)]]
```

## NAs - Select, See, and Eventually Remove
```{r, echo = TRUE, eval = TRUE}
x <- c(1, 2, NA, 4, NA, 6)
NAs <- is.na(x)
NAs

x[!NAs]
```

**Removing NAs on combination of several rows/cols**  
`complete.cases()` - Useful to keep only cases of database with only complete data accross multiple variables and subjects.  

```{r, echo = TRUE, eval = TRUE}
x <- c(1, 2, NA, 4)
y <- c("a", "b", "c", NA)
good <- complete.cases(x, y)
good
x[good]
y[good]
```

**To use in data frames**  
`good <- complete.cases(df)`, removes rows with NAs in accross all variables.  
`df[good, ][1:6, ]`, will contain only rows where there are no NAs  


### Sequences
```{r, echo = TRUE, eval = TRUE}
1:20
pi:10 # 10 isn't reached as it would be over
12:1
```

`seq()` args: `by`; `length`
```{r, echo = TRUE, eval = TRUE}
seq(1, 20)
seq(1, 5, by = 0.5)
seq(1, 5, length = 30)
```


```{r, echo = TRUE, eval = TRUE}
my_seq <- 1:10
1:length(my_seq)
seq(along.with = my_seq)
seq_along(my_seq)
```

**rep()**
```{r, echo = TRUE, eval = TRUE}
rep(0, times = 40)
rep(c(1, 2, 3), times = 10)
rep(c(1, 2, 3), each = 10)
```

### paste()
args:  
* `collapse` refers to separator of elements within a vector  
* `sep` refers to separator of elements from different vectors

```{r, echo = TRUE, eval = TRUE}
my_name <- c("My", "name", "is", "joao")
paste(my_name, collapse = " ")
```
```{r, echo = TRUE, eval = TRUE}
paste(c(1:3), c("X", "Y", "Z"), sep = "")
```
```{r, echo = TRUE, eval = TRUE}
paste(LETTERS, 1:4, sep = "-")
```

### Sample()
```{r, echo = TRUE, eval = TRUE}
y <- rnorm(100)
z <- rep(NA, 100)
my_data <- sample(c(y,z), 10)
my_data
```





###identical()
```{r, echoe = TRUE, eval = TRUE}
vect <- c(foo = 11, bar = 2, norf = NA)
vect2 <- c(11, 2, NA)
names(vect2) <- c("foo", "bar", "norf")
identical(vect, vect2)
```

###matrix()  
```{r, echoe = TRUE, eval = TRUE}
my_matrix <- matrix(data = 1:20, nrow = 4, ncol = 5)
```


###colnames()  
```{r, echoe = TRUE, eval = TRUE}
a <- c("John", 34, 65, 55, 45, 1)
b <- c("Mary", 44, 45, 68, 45, 2)
my_data <- data.frame(rbind(a, b)) # without data.frame, my_data would be a matrix and coerce all integers to char
cnames <- c("patient", "age", "weight", "bp", "rating", "test")
colnames(my_data) <- cnames
my_data
```

### LOGICAL OPERATORS  
`&` AND, recycles   
`&&` AND, does not recycle, eveluates first element of vector only   

`|` OR, recycles  
`||` OR, does not recycle, eveluates first element of vector only  

`&` has precedence over `|`  


### isTRUE()  
```{R echoe = TRUE, eval = TRUE}
isTRUE(5 > 4)
```

### identical()  
```{R echoe = TRUE, eval = TRUE}
identical('nip', 'nip')
```


### xor()   
Exclusive OR - TRUE if one (and only one) of args is TRUE  
```{R echoe = TRUE, eval = TRUE}
xor(TRUE, TRUE)
```

### which()  
Takes logical vector as argument and returns indices of the vector that are TRUE
```{R echoe = TRUE, eval = TRUE}
ints <- sample(10)
ints
which(ints > 7)
```

### any()  
Returns TRUE if one or more of the elements in the logical vector is TRUE
```{R echoe = TRUE, eval = TRUE}
ints <- sample(10)
ints
any(ints > 5)
```

### all()  
Returns TRUE if one or more of the elements in the logical vector is TRUE
```{R echoe = TRUE, eval = TRUE}
ints <- sample(10)
ints
all(ints > 0)
```

### Create a function that takes another function as argument  

``` {R echoe = TRUE, eval = FALSE}
  evaluate <- function(func, dat){
  result <- func(dat)
  result
}
```

### Anonimous Function
Function that is defined but not named
``` {R echoe = TRUE, eval = FALSE}
  evaluate <- function(func, dat){ # create function `evaluate`
  result <- func(dat)
  result
  }
  
evaluate(function(x){x+1}, 6)  # anonymous function inside `evaluate` function

# Example:
evaluate(function(x){x[1]}, c(8, 4, 0)) # Returns '8', the first element of vector 8,4,0
evaluate(function(x){x[length(x)]}, c(8, 4, 0)) # Returns '0', the last element of the vector
```  
Non-evaluated example  

`lapply(unique_vals, function(elem) elem[2])`

### Elipsis = ...  Allows an infinite number of arguments to be passed into a function.
When creating a function, all arguments created after an ellipses MUST have default values.
Normally  `...` is the last argument. But there are exceptions, such as in `paste()`

### Elipsis inside function  
```{R echo = TRUE, eval = FALSE}
telegram <- function(...){
  paste("START", ..., "STOP")
}
```

### Unpacking function args  
```{R echo = TRUE, eval = FALSE}
mad_libs <- function(...){
  # Do your argument unpacking here!
  args <- list(...)
  place <- args[["place"]]
  adjective <- args[["adjective"]]
  noun <- args[["noun"]]

  # Don't modify any code below this comment.
  # Notice the variables you'll need to create in order for the code below to
  # be functional!
  paste("News from", place, "today where", adjective, "students took to the streets in protest of the new", noun, "being installed on campus.")
}
```


### Create Binary Operator  
```{R echo = TRUE, eval = TRUE}
"%p%" <- function(left, right){
  paste(left, right, sep = " ")
}
"Hello" %p% "world!"
```

### lapply()
Takes a list and a function, and applies the funcion to each list member. On a datatable for example, it applies the function to each column.
`lapply(data, fun)` # data = datatable for example and fun some function such as class(). write fun name only, no parenthesis.  
lapply() returns a list - hence the _l_  

```{r, echoe = TRUE, eval = TRUE}
a <- c("John", 34, 65, 55, 45, 1)
b <- c("Mary", 44, 45, 68, 45, 2)
my_data <- data.frame(rbind(a, b)) # without data.frame, my_data would be a matrix and coerce all integers to char
cnames <- c("patient", "age", "weight", "bp", "rating", "test")
colnames(my_data) <- cnames
my_data$age <- c(100, 50)
my_data

lapply(my_data, class) # lapply returns list

# Another example
age <- as.data.frame(my_data[, 2])
age

lapply(age, mean)

```

### sapply()
sapply() uses lapply behind the scenes and then _simplifies_ the result transforming the list in vector, hence the _s_ fro simplify.

```{r, echoe = TRUE, eval = TRUE}
a <- c("John", 100, 65, 55, 45, 1)
b <- c("Mary", 50, 45, 68, 45, 2)
my_data <- data.frame(rbind(a, b)) # without data.frame, my_data would be a matrix and coerce all integers to char
cnames <- c("patient", "age", "weight", "bp", "rating", "test")
colnames(my_data) <- cnames
my_data$age <- c(100, 50)
my_data

sapply(my_data, class) # sapply resturns vector

# Another example

age <- as.data.frame(my_data[, 2])
age
sapply(age, mean)
```

### Extract percentage. See _Binary data makes it easy to compute percentage_ below


```{r, echoe = TRUE, eval = TRUE}
a <- c(1, 0, 1)
b <- c(1, 1, 0)
c <- c(1, 1, 1)
d <- c(1, 0, 0)
e <- c(1, 1, 0)
my_data <- data.frame(rbind(a, b, c, d, e)) # without data.frame, my_data would be a matrix and coerce all integers to char
cnames <- c("Red", "Blue", "Yellow")
colnames(my_data) <- cnames
my_data

sapply(my_data, mean)

# 1 is 100%, 0.6 is 60%, 0.4 is 40%. % of times that each color is used (value = 1)
```


**sapply output**  

>In general, if the result is a list where every element is of length one, then sapply() returns a vector. If the result is a list where every element is a vector of the same length (> 1), sapply() returns a
matrix. If sapply() can't figure things out, then it just returns a list, no different from what lapply()
would give you. _Source: swirl() package._  

### Binary data makes it easy to compute percentage

>Perhaps it's more informative to find the proportion of flags (out of 194) containing each color. Since > each column is just a bunch of 1s and 0s, the arithmetic mean of each column will give us the proportion > of 1s. (If it's not clear why, think of a simpler situation where you have three 1s and two 0s -- (1 + 1 > + 1 + 0 + 0)/5 = 3/5 = 0.6). _Source: swirl() package_


### range()  
Returns the min and max values of a numeric vector (first argument)

```{r, echoe = TRUE, eval = TRUE}
a <- c(0, 1, 0)
b <- c(30, 1, 0)
c <- c(1, 0, 15)
d <- c(1, 5, 0)
e <- c(1, 10, 0)
my_data <- data.frame(rbind(a, b, c, d, e)) # without data.frame, my_data would be a matrix and coerce all integers to char
cnames <- c("Star", "Square", "Circle")
colnames(my_data) <- cnames
my_data

sapply(my_data, range)

# returns the min (row 1) and max (row 2) values for each col.
```

### list.files()  
Returns character vector of the names of files or directories in the original directory.   

### Conditional Subsetting with [[]]
andy[which(andy$Day == 30), "Weight"]
andy[which(andy[,"Day"] == 30), "Weight"]


### do.call()  

> do.call lets you specify a function and then passes a list as if each element of the list were an argument to the function. _Source: https://github.com/derekfranks/practice_assignment/blob/master/practice_assignment.rmd_

do.call(function_you_want_to_use, list_of_arguments)  


